Natural variations in water discharge quantity have an inconsistent impact on sediment transport capacity in subglacial channels

### 2023-10-09 Delaneyetal2023b.tex
Appendix done

Here three kinds of drainage regimes are examined, namely, subaerial channels, steady-state R-channels \cite{rothlisberger1972}, and pipe-flow (i.e. R-channels which do not have time to adjust their size to discharge conditions).
Sediment transport capacity is scaled with respect to a given water discharge for three different transport formulas: MPM \cite{meyer1948}, EH \cite{engelund1967}, and Bagnold \cite{bagnold1980}; additionally width-integrated shear stress is assessed, as it is  used as proxy for sediment transport throughout the manuscript.

A few basic hydraulic relations are used in the following.
The Darcy-Weisbach equation can be stated as
\begin{equation}
  \label{eq:DW}
  \Psi \propto f\frac{Q^2}{D_h S^2},
\end{equation}
with $\Psi$ the head gradient and friction factor $f$.
The shear stress and stream power are, respectively,
\begin{equation}
    \label{eq:tau-omega}
  \tau \propto f v^2 = f \left(\frac{Q}{S}\right)^2, \quad  \Omega \propto \Psi Q.
\end{equation}
%

Sediment discharge is given by the MPM, EH and Bagnold formulations as, respectively,
\begin{equation}
  Q_s \propto w_c\, \tau^{3/2}, \quad Q_s \propto w_c\, \tau^{5/2}, \quad Q_s \propto w_c \left(\frac{\Omega}{w_c}\right)^{3/2} H^{-2/3}
\end{equation}
for conditions well above the transport onset threshold.

The  subaerial channels is assumed to have a width $w_c$ much greater than its depth $H$, such that $D_h\approx 4h$, and to have a constant head gradient ($\Psi$) given by the topography.
Further, it is  assumed that its width can be approximated by a relation $w_c \propto Q^\alpha$ (Equation~\ref{eq:wcf})with $\alpha\in [0,1]$; typically $\alpha \approx 1/2$ (a so-called regime channel).
End members $\alpha=0$ or $\alpha=1$ correspond a  subaerial channels of constant width (a slot canyon) or depth (no natural equivalent), respectively.
%
For a steady state R-channel, it is assumed that  $\Psi$ is constant (approximated by the gradient of the Shreve \citeyear{shreve1972} potential) and that $S$ adjusts such that it is in steady state with $\Psi$ and $Q$.  %Conversely, pipe flow has $S$ fixed and $\Psi$ adjusts to satisfy the given $Q$.
%
Pipe flow is approximated when a R-channel is subjected to rapid discharge variations such that the channel does not adjust its size significantly on the time scale of the discharge variations.
In this case, it is assumed that the cross-sectional area $S$ is fixed, and $\Psi$ adjusts to the specified $Q$.
For both R-channel and pipe flow, it is assumed assume that $D_h \propto S^{1/2}$.

@@

Table~\ref{tab:Qs2} shows that the proxy $w_c\tau$ as well as the MPM, EH and Bagnold sediment transport formulas give roughly a similar scaling of $Q_s$ on hydraulic gradient $\Psi$ or $S$ and on water discharge $Q$.
Remarkably, the Bagnold formula has a negative exponent for $f$ in all but the pipe-flow regime, which seems rather unexpected.
The total transport formula EH gives a slightly stronger dependence on all variables as should be expected due to the larger exponent on $\tau$ of $\frac{5}{2}$ versus  $\frac{3}{2}$  for the others.
Albeit, the sediment transport response in pipe-flow for the Bagnold case is close to EH.
Conversely, the sediment transport proxy $t_{tg},t_{tf}$ used within this publication scales only as $Q^2$ for pipe-flow, whereas the others scale at least with $Q^3$.
The exponent for the width scaling $\alpha$ only impacts relationship between sediment transport and water discharge in the EH relation in any meaningful way.
However, for the value of $\alpha$ around $\frac{1}{2}$,  suits braided proglacial stream and indeed most  subaerial channels.
Furthermore, when $\alpha=1$, the exponent on $Q$ is only slightly above the other values of $\alpha$.

The sediment discharge in the steady state R-channel scales very similarly to the  subaerial channel case for all relations and virtually identically for the $\alpha=0.5$ case.  However, note that the head gradients $\Psi$ are likely higher for comparable $Q$ in a ice-sheet marginal or alpine glacier setting than in a  subaerial channel, as is well described in \citeA{alley1997}, and thus transport rates may still be much higher in a steady state R-channel.

Furthermore, R-channels will rarely operate in steady state as variations in discharge, in particular on the diurnal timescale or the timescale of severe rain or melting events, are too fast for such a channel to reach a steady state.
In such cases they operate more like a pipe of fixed cross-section \cite<e.g.>{gimbert2016}.
Table~\ref{tab:Qs2} shows that in such a situation, the sediment transport scales much more severely with discharge, with the exponent on $Q$ being likely between $3$ and $5$, compared to other two regimes when that exponent is at most $1.3$ (Table~\ref{tab:Qs2}).
Thus fluctuations of discharge on short timescales (on the order of a day) have the potential to cause very high sediment transport conditions.

\begin{table}
  \caption{Relations for hydraulic variables for the three drainage regimes:  subaerial channels, R-channel and pipe-flow.  Darcy-Weisbach equation is abbreviated with ``D-W'', and stream power with ``Stream p.''. }
  \label{tab:eqs1}
  \small
\begin{tabular}{llllll}
  Regime & Fixed & \makecell{Determined via D-W
  \\Equation~\ref{eq:DW}} & Additional relations & Shear stress & Stream p.\\
 &  &  &  & \(\tau \propto\) & \(\Omega \propto\)\\
\hline
  Subaerial & \(\Psi\) & \(H \propto f^{1/3}\, Q^{2/3-2\alpha/3} \, \Psi^{-1/3}\) & 
\makecell{\(w_c\,\propto Q^\alpha\) \\ \(S=w_cH\) \\ \(D_h=H\)} & 
\(f^{1/3} Q^{2/3-2\alpha/3}  \Psi^{2/3}\) & \(Q \Psi\)\\
R-ch. & \(\Psi\) & \(S\, \propto f^{2/5}\, Q^{4/5} \, \Psi^{-2/5}\) & \(D_h\propto w_c \propto H \propto S^{1/2}\) & \(f^{1/5} Q^{2/5} \, \Psi^{4/5}\) & \(Q \Psi\)\\
Pipe & \(S\) & \(\Psi \propto f \, Q^2\, S^{-5/2}\) & \(D_h\propto w_c \propto H \propto S^{1/2}\) & \(f Q^2 S^{-2}\) & \(f Q^3 S^{-5/2}\)\\
\end{tabular}
\end{table}

\begin{table}
  \caption{Sediment transport proxy ($w_c\tau$) and rates for the three considered different transport formulas: MPM \cite{meyer1948}, EH \cite{engelund1967}, and Bagnold \cite{bagnold1980}.
    }
  \label{tab:Qs}
\begin{tabular}{lllll}
 & Width*\(\tau\) & MPM & EH & Bagnold\\
 & \(w_c\, \tau\) & \(Q_s \propto w_c\, \tau^{3/2}\) & \(Q_s \propto w_c\, \tau^{5/2}\) & \(Q_s \propto w_c^{-1/2}\, \Omega^{3/2} H^{-2/3}\)\\
\hline
Subaerial  & \(f^{1/3}\, Q^{2/3+\alpha/3}\,  \Psi^{2/3}\) & \(f^{1/2}\, Q \, \Psi\) & \(f^{5/6}\, Q^{5/3 - 2\alpha/3} \, \Psi^{5/3}\) & \(f^{-2/9}\, Q^{19/18-\alpha/18} \, \Psi^{31/18}\)\\
R-ch. & \(f^{2/5}\, Q^{4/5} \, \Psi^{3/5}\) & \(f^{1/2}\, Q \, \Psi\) & \(f^{7/10}\, Q^{7/5}\, \Psi^{9/5}\) & \(f^{-7/30}\, Q^{31/30}\, \Psi^{26/15}\)\\
Pipe & \(f \, Q^2 \, S^{-1}\) & \(f^{3/2}\, Q^3 \, S^{-5/2}\) & \(f^{5/2}\, Q^5\, S^{-9/2}\) & \(f^{3/2} \, Q^{9/2} \, S^{-14/3}\)\\
\end{tabular}
\end{table}

\begin{table}
  \caption{As Table~\ref{tab:Qs} but with the fractional exponents stated as rounded decimal numbers.  For the subaerial regime the exponents are for displayed for the likely width-exponent $\alpha=1$, as well as its end members 0 (slot canyon) and 1 (only width increases and not depth).
    }
  \label{tab:Qs2}
\begin{tabular}{lllll}
 & Width*\(\tau\) & MPM & EH & Bagnold\\
 & \(w_c\, \tau\) & \(Q_s \propto w_c\, \tau^{3/2}\) & \(Q_s \propto w_c\, \tau^{5/2}\) & \(Q_s \propto w_c^{-1/2}\, \Omega^{3/2} H^{-2/3}\)\\
\hline
Subaerial (\(\alpha=0\)) & \(f^{0.7}\, Q^{0.7}\,  \Psi^{0.7}\) & \(f^{0.5}\, Q \,\,\,\, \Psi\) & \(f^{0.8}\, Q^{1.7} \, \Psi^{1.7}\) & \(f^{-0.2}\, Q^{1.1} \, \Psi^{1.7}\)\\
Subaerial (\(\alpha=0.5\)) & \(f^{0.7}\, Q^{0.8}\,  \Psi^{0.7}\) & \(f^{0.5}\, Q \,\,\,\, \Psi\) & \(f^{0.8}\, Q^{1.3} \, \Psi^{1.7}\) & \(f^{-0.2}\, Q^{1.0} \, \Psi^{1.7}\)\\
Subaerial (\(\alpha=1\)) & \(f^{0.7}\, Q^{1.0}\,  \Psi^{0.7}\) & \(f^{0.5}\, Q \,\,\,\, \Psi\) & \(f^{0.8}\, Q^{0.7} \, \Psi^{1.7}\) & \(f^{-0.2}\, Q^{1.0} \, \Psi^{1.7}\)\\[3pt]
R-ch. & \(f^{0.4}\, Q^{0.8} \, \Psi^{0.6}\) & \(f^{0.5}\, Q \,\,\,\, \Psi\) & \(f^{0.7}\, Q^{1.4}\, \Psi^{1.8}\) & \(f^{-0.2}\, Q^{1.0} \, \Psi^{1.7}\)\\
Pipe & \(f \,\quad Q^{2\phantom{.0}} \, S^{-1}\) & \(f^{1.5}\, Q^3 \, S^{-2.5}\) & \(f^{2.5}\, Q^{5\phantom{.0}}\, S^{-4.5}\) & \(f^{1.5} \,\,\,\,\, Q^{4.5} \, S^{-4.7}\)\\
\end{tabular}
\end{table}

>>>>

Here three kinds of drainage regimes are examined, namely, subaerial channels, steady-state R-channels \cite{rothlisberger1972}, and pipe-flow (i.e. R-channels which do not have time to adjust their size to discharge conditions).
Here we show how sediment transport capacity scales with respect to a given water discharge and hydraulic gradient for three different transport formulas: MPM \cite{meyer1948}, EH \cite{engelund1967}, and Bagnold \cite{bagnold1980}; additionally width-integrated shear stress is assessed, as used as proxy for sediment transport throughout the manuscript.

A few basic hydraulic relations are used in the following.
The Darcy-Weisbach equation can be stated as
\begin{equation}
  \label{eq:DW}
  \Psi \propto f\frac{Q^2}{D_h S^2},
\end{equation}
with $\Psi = \Delta h / l$ the head gradient and friction factor $f$.
The shear stress and stream power are, respectively,
\begin{equation}
    \label{eq:tau-omega}
  \tau \propto f v^2 = f \left(\frac{Q}{S}\right)^2, \quad  \Omega \propto \Psi Q.
\end{equation}
%

Sediment discharge is given by the MPM, EH and Bagnold formulations as
\begin{equation}
  Q_s \propto w_c\, \tau^{3/2}, \quad Q_s \propto w_c\, \tau^{5/2}, \quad Q_s \propto w_c \left(\frac{\Omega}{w_c}\right)^{3/2} H^{-2/3},
\end{equation}
respectively, for conditions well above the transport onset threshold.

The  subaerial channel is assumed to have a width $w_c$ much greater than its depth $H$, such that $D_h\approx 4h$, and to have a constant head gradient ($\Psi$) given by the topography.
Further, it is  assumed that its width can be approximated by a relation $w_c \propto Q^\alpha$ (Equation~\ref{eq:wcf})with $\alpha\in [0,1]$; typically $\alpha \approx 1/2$ (a so-called regime channel).
End members $\alpha=0$ or $\alpha=1$ correspond a  subaerial channels of constant width (a slot canyon) or depth (no natural equivalent), respectively.
%
For a steady state R-channel, it is assumed that  $\Psi$ is constant (approximated by the gradient of the Shreve \citeyear{shreve1972} potential \footnote{Note that the R-channel model presented above (Eq.~\eqref{eq:dS_dt}--\eqref{eq:dh2wc}), calculates $\Psi$ from the time evolving $S_g$ (Eq.~\eqref{eq:dS_dt}) via the Darcy-Weisbach equation~\eqref{eq:dh} and thus no Shreve approximation is then needed.}) and that $S$ adjusts such that it is in steady state with $\Psi$ and $Q$.
% Conversely, pipe flow has $S$ fixed and $\Psi$ adjusts to satisfy the given $Q$.
%
Pipe flow like conditions occur when a R-channel is subjected to rapid discharge variations such that the channel cannot adjust its size significantly on the time scale of the discharge variations.
In this case, it is assumed that the cross-sectional area $S$ is fixed, and $\Psi$ adjusts to the specified $Q$.
For both R-channel and pipe flow, it is assumed assume that $D_h \propto S^{1/2}$.

@@

Table~\ref{tab:Qs2} shows how the proxy $w_c\tau$ as well as $Q_s$ of the MPM, EH and Bagnold sediment transport formulas scale with respect to $Q$, $\Psi$ or $S$ for the three different regimes and for different $\alpha$.
Remarkably, the Bagnold formula has a negative exponent for $f$ in all but the pipe-flow regime, which seems rather unexpected.
The total transport formula EH gives a slightly stronger dependence on all variables as should be expected due to the larger exponent on $\tau$ of $\frac{5}{2}$ versus  $\frac{3}{2}$  for the others.
Albeit, the sediment transport response in pipe-flow for the Bagnold case is close to EH.
Conversely, the sediment transport proxy $w_g\,\tau_{g}, w_f\,\tau_{f}$ used within this publication scales only as $Q^2$ for pipe-flow, whereas the others scale at least with $Q^3$.
The exponent for the width scaling $\alpha$ only impacts relationship between sediment transport and water discharge in the EH relation in any meaningful way.
However, for the value of $\alpha$ around $\frac{1}{2}$ (a value appropriate for most streams), the exponent on $Q$ for EH is only slightly above what the other transport relations give.

The sediment discharge in the steady state R-channel scales very similarly to the  subaerial channel case for all relations and virtually identically for the $\alpha=0.5$ case.  However, note that the head gradients $\Psi$ are likely higher for comparable $Q$ in an ice-sheet marginal or alpine glacier setting than in a  subaerial channel, as is well described in \citeA{alley1997}, and thus transport rates may still be much higher in a steady state R-channel.

Furthermore, R-channels will rarely operate in steady state as variations in discharge, in particular on the diurnal timescale or the timescale of severe rain or melting events, are too fast for such a channel to reach a steady state.
In such cases they operate more like a pipe of fixed cross-section \cite<e.g.>{gimbert2016}.
Table~\ref{tab:Qs2} shows that in such a situation, the sediment transport scales much more severely with discharge, with the exponent on $Q$ being likely between $3$ and $5$, compared to other two regimes when that exponent is at most $1.3$ (Table~\ref{tab:Qs2}).
Thus fluctuations of discharge on short timescales (on the order of a day) have the potential to cause conditions with very high sediment transport capacities.

\begin{table}
  \caption{Relations for hydraulic variables for the three drainage regimes:  subaerial channels, R-channel and pipe-flow.  Darcy-Weisbach equation is abbreviated with ``D-W'', and stream power with ``Stream p.''. }
  \label{tab:eqs1}
\begin{tabular}{llllll}
  Regime & Fixed & Determined via D-W
   & Additional relations & Shear stress & Stream p.\\
& & (Equation~\ref{eq:DW})  &  &  \(\tau \propto\) & \(\Omega \propto\)\\
\hline
  Subaerial & \(\Psi\) & \(H \propto f^{1/3}\, Q^{2/3-2\alpha/3} \, \Psi^{-1/3}\) & \makecell{\(w_c\,\propto Q^\alpha\) \\ \(S=w_cH\) \\ \(D_h\propto H\)} & \(f^{1/3} Q^{2/3-2\alpha/3}  \Psi^{2/3}\) & \(Q\, \Psi\)\\
  R-channel & \(\Psi\) & \(S\, \propto f^{2/5}\, Q^{4/5} \, \Psi^{-2/5}\) & \(D_h\propto w_c \propto H \propto S^{1/2}\) & \(f^{1/5} Q^{2/5} \, \Psi^{4/5}\) & \(Q\, \Psi\)\\
Pipe & \(S\) & \(\Psi \propto f \, Q^2\, S^{-5/2}\) & \(D_h\propto w_c \propto H \propto S^{1/2}\) & \(f Q^2 S^{-2}\) & \(f\, Q^3 S^{-5/2}\)\\
\end{tabular}
\end{table}

\begin{table}
  \caption{Sediment transport proxy ($w_c\tau$) and rates for the three considered different transport formulas: MPM \cite{meyer1948}, EH \cite{engelund1967}, and Bagnold \cite{bagnold1980}.
    }
  \label{tab:Qs}
\begin{tabular}{lllll}
 & Width \(\times \,\, \tau\) & MPM & EH & Bagnold\\
 & \(w_c\, \tau\) & \(Q_s \propto w_c\, \tau^{3/2}\) & \(Q_s \propto w_c\, \tau^{5/2}\) & \(Q_s \propto w_c^{-1/2}\, \Omega^{3/2} H^{-2/3}\)\\
\hline
Subaerial  & \(f^{1/3}\, Q^{2/3+\alpha/3}\,  \Psi^{2/3}\) & \(f^{1/2}\, Q \, \Psi\) & \(f^{5/6}\, Q^{5/3 - 2\alpha/3} \, \Psi^{5/3}\) & \(f^{-2/9}\, Q^{19/18-\alpha/18} \, \Psi^{31/18}\)\\
R-channel & \(f^{2/5}\, Q^{4/5} \, \Psi^{3/5}\) & \(f^{1/2}\, Q \, \Psi\) & \(f^{7/10}\, Q^{7/5}\, \Psi^{9/5}\) & \(f^{-7/30}\, Q^{31/30}\, \Psi^{26/15}\)\\
Pipe & \(f \, Q^2 \, S^{-1}\) & \(f^{3/2}\, Q^3 \, S^{-5/2}\) & \(f^{5/2}\, Q^5\, S^{-9/2}\) & \(f^{3/2} \, Q^{9/2} \, S^{-14/3}\)\\
\end{tabular}
\end{table}

\begin{table}
  \caption{As Table~\ref{tab:Qs} but with the fractional exponents stated as rounded decimal numbers.  For the subaerial regime the exponents are for displayed for the likely width-exponent $\alpha=1$, as well as its end members 0 (slot canyon) and 1 (only width increases and not depth).
    }
  \label{tab:Qs2}
\begin{tabular}{lllll}
 & Width \(\times \,\, \tau\) & MPM & EH & Bagnold\\
 & \(w_c\, \tau\) & \(Q_s \propto w_c\, \tau^{3/2}\) & \(Q_s \propto w_c\, \tau^{5/2}\) & \(Q_s \propto w_c^{-1/2}\, \Omega^{3/2} H^{-2/3}\)\\
\hline
Subaerial (\(\alpha=0\)) & \(f^{0.7}\, Q^{0.7}\,  \Psi^{0.7}\) & \(f^{0.5}\, Q \,\,\,\, \Psi\) & \(f^{0.8}\, Q^{1.7} \, \Psi^{1.7}\) & \(f^{-0.2}\, Q^{1.1} \, \Psi^{1.7}\)\\
Subaerial (\(\alpha=0.5\)) & \(f^{0.7}\, Q^{0.8}\,  \Psi^{0.7}\) & \(f^{0.5}\, Q \,\,\,\, \Psi\) & \(f^{0.8}\, Q^{1.3} \, \Psi^{1.7}\) & \(f^{-0.2}\, Q^{1.0} \, \Psi^{1.7}\)\\
Subaerial (\(\alpha=1\)) & \(f^{0.7}\, Q^{1.0}\,  \Psi^{0.7}\) & \(f^{0.5}\, Q \,\,\,\, \Psi\) & \(f^{0.8}\, Q^{0.7} \, \Psi^{1.7}\) & \(f^{-0.2}\, Q^{1.0} \, \Psi^{1.7}\)\\[3pt]
R-channel & \(f^{0.4}\, Q^{0.8} \, \Psi^{0.6}\) & \(f^{0.5}\, Q \,\,\,\, \Psi\) & \(f^{0.7}\, Q^{1.4}\, \Psi^{1.8}\) & \(f^{-0.2}\, Q^{1.0} \, \Psi^{1.7}\)\\
Pipe & \(f \,\quad Q^{2\phantom{.0}} \, S^{-1}\) & \(f^{1.5}\, Q^3 \, S^{-2.5}\) & \(f^{2.5}\, Q^{5\phantom{.0}}\, S^{-4.5}\) & \(f^{1.5} \,\,\,\,\, Q^{4.5} \, S^{-4.7}\)\\
\end{tabular}
\end{table}
